Towards a Verified Bzip2

### 2019-10-07 draft_paper/draft_paper.tex
Add CPP conference
Change citation style to author/year
Small cleanups
Inline all the listings

\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2020}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

@@

As a part of the proof of \Cref{thm:unbwt_correct}, we proved two
other noteworthy results that have not, to our knowledge, been proved
before in Coq. We implemented a least-significant-digit (LSD)
radixsort and proved its correctness,
\begin{restatable*}[radixsort\_correct]{theorem}{radixsortcorrect}
  \label{thm:radixsort_correct}%
  For all matrices $m$ with $n$ columns,
  \begin{equation*}
    \var{Sorted}~(\var{radixsort}~m~n) \land \var{StablePerm}~(\var{radixsort}~m~n)~m
  \end{equation*}
\end{restatable*}
and we proved that any two stable sorts are interchangeable,
\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
  \label{thm:stablesort_unique}
  For all functions $f, g: \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    \var{StableSort}~f \implies \var{StableSort}~g \implies
    (\forall l,~f~l = g~l)
  \end{equation*}
\end{restatable*}
This in turn required developing suitable definitions for
\var{Sorted}, \var{Permutation}, and \var{StablePerm}.

@@

We represent a total, decidable preorder on type with the
\var{Preord} typeclass in Coq:
\lstinputlisting[firstline=9, lastline=14]{../theories/BWT/Sorting/Ord.v}
We do not explicitly require reflexivity because it follows from
totality. Given a \var{Preord} instance for a type $A$, we will be
able to sort $\var{list}~A$ as shown in \cref{sec:sort}.

@@

We say that we have a total, decidable \textit{order} when we have
that for all $x$ and $y$,
\begin{equation*}
  x \equiv y \implies x = y
\end{equation*}
that is, the induced equivalence relation of the preorder ($\equiv$)
implies Leibniz equality ($=$). We use the \var{Ord} class in Coq
for this:
\lstinputlisting[firstline=175, lastline=176, widthgobble=2]{../theories/BWT/Sorting/Ord.v}
For the BWT, we require an \var{Ord} instance for $A$, but
elsewhere, such as in sorting, \var{Preord} will suffice.

@@

First, we define the left and right rotations of a list. \var{lrot}
moves the first element of a list to the end.
\lstinputlisting[widthgobble=2, firstline=24, lastline=28]{../theories/BWT/Rotation/Rotation.v}
\var{rrot} moves the last element of a list to the front, and
relies on \var{init}, which drops the last element of a list.
\lstinputlisting[widthgobble=2, firstline=18, lastline=22]{../theories/BWT/Lib/List.v}
\lstinputlisting[widthgobble=2, firstline=17, lastline=21]{../theories/BWT/Rotation/Rotation.v}
Note that in order to ensure that $\var{last}$ is total, it takes
an extra argument to return in the case of an empty list.

Now we produce the rotation matrix by repeatedly applying \var{lrot}.
We will use \var{iter}, which takes a function $f : A \to A$, a number
$n$, and an initial value $z$, and returns the list of intermediate
values obtained by applying $f$ to $z$ $n$ times. Or,
\begin{equation}
  \var{iter}~f~n~z = [z,~f~z,~f~(f~z),~\ldots,~f^{n-1} z]
  \label{eq:iter}
\end{equation}
where $f^n$ denotes repeated function application.
\lstinputlisting[
  widthgobble=2,
  firstline=14,
  lastline=18
]{../theories/BWT/Lib/Iterate.v}

We can state \Cref{eq:iter} as a theorem by defining the
function \var{rep} for repeated function application.
\lstinputlisting[
  widthgobble=2, firstline=8, lastline=12,
]{../theories/BWT/Lib/Repeat.v}
In proofs, we will still use the notation $f^i z$ for
$\var{rep}~f~i~z$. We will also use that notation $l_i$ for the $i$th
element of a list. In Coq, that is given by $\var{nth}~i~l~d$, where
$d:A$ is a default value in case $i \ge \var{length}~l$.

@@

Now we can define \var{rots} to produce the rotation matrix.
\lstinputlisting[
  widthgobble=2,
  firstline=17,
  lastline=18,
]{../theories/BWT/Rotation/Rots.v}

\var{bwl} produces the transformed string by taking the last column of
the sorted rotation matrix. \var{lexsort} sorts the rotation matrix
using insertion sort (\cref{appendix:insertion_sort}) and the induced
lexicographic ordering(\cref{sec:lex_ord}). By
\Cref{thm:stablesort_ord} though, \var{lexsort} can be implemented
using any sorting algorithm.
\lstinputlisting[
  widthgobble=2,
  firstline=31,
  lastline=35
]{../theories/BWT/BurrowsWheeler.v}

\var{bwn} returns the index in the sorted rotation matrix where the
original string appears. It relies on \var{findIndex}, which is
defined with respect to some decidable equivalence relation.
\lstinputlisting[
  widthgobble=2,
  firstline=12,
  lastline=19,
]{../theories/BWT/Lib/FindIndex.v}

When we use \var{findIndex} in \var{bwn}, we are using the
induced equivalence relation from the \var{Ord} instance, so it
finds the row that is Leibniz equal to $l$.
\lstinputlisting[
  widthgobble=2,
  firstline=75,
  lastline=75,
]{../theories/BWT/BurrowsWheeler.v}

@@

We implement \var{recreate} as follows, introducing a parameter $j$ so
that we can recursively recreate the first $j$ columns of the matrix,
one column at a time. Here $l$ is meant to be the output of the BWT,
the last column of the sorted rotation matrix.
\lstinputlisting[
  widthgobble=2, firstline=156, lastline=160,
]{../theories/BWT/BurrowsWheeler.v}
This definition relies on \var{prepend\_col} and \var{zipWith}, defined below.
\lstinputlisting[
  widthgobble=2, firstline=47, lastline=48,
]{../theories/BWT/Columns.v}
\lstinputlisting[
  widthgobble=2, firstline=16, lastline=20,
]{../theories/BWT/Lib/ZipWith.v}

Now we can redefine \var{unbwt} as
\lstinputlisting[
  widthgobble=2,
  firstline=197,
  lastline=198,
]{../theories/BWT/BurrowsWheeler.v}

Since we recreate the matrix column by column, we can define an
invariant using $\var{cols}~j$, which takes the first $j$ columns of a matrix.
\lstinputlisting[
  widthgobble=2,
  firstline=19,
  lastline=20,
]{../theories/BWT/Columns.v}

@@

This leads to the following definition of the property
$\var{PermFun}~n~p$, which holds when $p : \var{list}~\mathbb{N}$
represents a permutation function on lists of length $n$:
\lstinputlisting[
  firstline=14, lastline=15,
]{../theories/BWT/Sorting/PermFun.v}
Then, we can define \var{apply}, which permutes a list according to a
given permutation:
\lstinputlisting[
  widthgobble=2, firstline=161, lastline=165,
]{../theories/BWT/Sorting/PermFun.v}
And finally, we can define \var{PermutationEx}, which holds between
two lists when they are permutations of each other.
\lstinputlisting[
  widthgobble=2, firstline=521, lastline=522,
]{../theories/BWT/Sorting/PermFun.v}

We prove \var{PermutationEx} equivalent to \var{Permutation} in
\Cref{thm:permutationex_iff}.

For completeness, we will briefly mention a third definition of a
permutation relation that says that two lists are permutations of one
another when any element occurs an equal number of times in each:
\lstinputlisting[
  widthgobble=2, firstline=14, lastline=15,
]{../theories/BWT/Sorting/PermutationCount.v}
\var{count\_occ} counts the number of occurrences of a given variable
using a decision procedure for Leibniz equality. Here that decision
procedure is \texttt{equiv\_dec} from an \var{EqDec} instance where we
require the equivalence relation itself by Leibniz equality.

@@

Based on this, we define a stronger version of the \var{PermFun}
property, \var{StablePermFun}. However, while $\var{PermFun}~n~p$
means that $p$ represents a permutation function on lists of length
$n$, we need to specify the list with respect to which $p$ is stable,
so we say $\var{StablePermFun}~l~p$ for some list $l$.
\lstinputlisting[
  widthgobble=2, firstline=568, lastline=572,
]{../theories/BWT/Sorting/PermFun.v}
As above, we can also define this
property using the inverse permutation:
\lstinputlisting[
  widthgobble=2, firstline=574, lastline=578,
]{../theories/BWT/Sorting/PermFun.v}
As usual, we will prove these
definitions equivalent. First, though, we need to define \var{image}
and \var{preimage}.
\lstinputlisting[
  widthgobble=2, firstline=81, lastline=82,
]{../theories/BWT/Sorting/PermFun.v}
We will not prove
it here, but $\var{preimage}~p$ and $\var{image}~p$ are both of
bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$ and are
inverses of each other. We also have the following two theorems:
\begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
  For all lists $l$, permutations $p$, and indices $i$,
  \begin{align*}
    (\var{apply}~p~l)_{p^{-1}(i)} = l_i && l_{p(i)} = (\var{apply}~p~l)_i
  \end{align*}
\end{theorem}
\begin{proof}
  The first equation (\var{nth\_preimage\_apply}) holds because:
  \begin{align*}
       (\var{apply}~p~l)_{p^{-1}(i)}
    &= (\var{map}~(\lambda j.~l_j)~p)_{\var{findIndex}~i~p} \\
    &= l_{p_{(\var{findIndex}~i~p)}} \\
    &= l_i
  \end{align*}
  and the second follows from this:
  \begin{equation*}
      (\var{apply}~p~l)_i
    = (\var{apply}~p~l)_{p^{-1}(p(i))}
    = l_{p(i)}
    \qedhere
  \end{equation*}
\end{proof}

@@

Now, we can define \var{StablePermEx} to describe when two lists are
stable permutations of each other:
\lstinputlisting[
  widthgobble=2, firstline=618, lastline=619,
]{../theories/BWT/Sorting/PermFun.v}

While \var{StablePermEx} corresponds to \var{PermutationEx}, how can
we modify \var{Permutation} to create an inductive relation for stable
lists? Intuitively, the problematic inference rule is
\var{Permutation\_swap}: we can only swap two elements and preserve
stability if they are not equivalent. With this fix, we have \var{StablePermInd}:
\begin{gather*}
  \infer[\var{StablePermInd\_nil}]{\var{StablePermInd}~[]~[]}{}
  \\ \\
  \infer[\var{StablePermInd\_skip}]{\var{StablePermInd}~(x :: l)~(x :: l')}{%
    \var{StablePermInd}~l~l'
  }
  \\ \\
  \infer[\var{StablePermInd\_swap}]
        {\var{StablePermInd}~(x :: y :: l)~(y :: x :: l)}
        {x \not\equiv y}
  \\ \\
  \infer[\var{StablePermInd\_trans}]{\var{StablePermInd}~l~l''}{
    \var{StablePermInd}~l~l' & \var{StablePermInd}~l'~l''
  }
\end{gather*}

Unfortunately, neither of these two definitions will be particularly
useful for proving the theorems we would like to prove with them. The
best definition for stability does not come by analogy with
\var{Permutation}. The intuition is that for any two lists that are
stable permutations of each other, if we filter each list by
equivalence with any element, the resulting lists should be identical.
This is because they must have the same elements, since the original
lists are permutations of one another, and those elements must be in
the same order, since the original lists are specifically stable
permutations of one another \cite{leroy_2018}.
\lstinputlisting[
  widthgobble=2, firstline=21, lastline=22,
]{../theories/BWT/Sorting/StablePerm.v}

We prove equivalence between these three definitions in
\cref{appendix:stableperm_def_eq}.

@@

If we define \var{Sort} and \var{StableSort} as we did
\cref{sec:sorted_perm_stable}, we can restate the theorem slightly
more elegantly.
\lstinputlisting[
  widthgobble=2, firstline=15, lastline=19,
]{../theories/BWT/Sorting/Sort.v}

@@

Taken together, we can define prefix lexicographic orderings on lists,
where the lexicographic ordering is only allowed to compare up to the
first $n$ elements of two lists. We define the relation $\le_n$ on
$\var{list}~A$ such that for all $xs, ys : \var{list}~A$,
\begin{equation*}
  xs \le_n ys \coloneqq (\var{firstn}~n~xs) \lexle (\var{firstn}~n~ys)
\end{equation*}
For convenience, we define $\var{PrefixSorted}~n$ to hold when a list
of lists is sorted if you only look at the first $n$ elements of each:
\lstinputlisting[
  widthgobble=2, firstline=326, lastline=327,
]{../theories/BWT/Sorting/Key.v}

@@

In a functional setting, we can define an LSD radixsort elegantly by
repeatedly moving the last column to the front of the matrix and
sorting on the first column. For a matrix with $n$ columns,
\begin{equation*}
  \var{radixsort} = (\var{hdsort} \comp \var{map}~\var{rrot})^n
\end{equation*}
Or, in Coq:
\lstinputlisting[
  widthgobble=2, firstline=27, lastline=28,
]{../theories/BWT/Sorting/RadixSort.v}
Here \var{hdsort} is just an insertion sort using $\le_1$ ordering.

@@

\lstinputlisting[
  widthgobble=2, firstline=15, lastline=26,
]{../theories/BWT/Sorting/InsertionSort.v}

@@

We know we will eventually need to reason inductively about
permutation functions, so we would like to define the relationship
between applying a permutation function $i :: p$ and applying $p$.
Conceptually, the first element of $\var{apply}~(i::p)~l$ will be
$l_i$. The rest will be the result of applying some new permutation
$p'$, where everything greater than $i$ has been shifted down by one,
to some new list $l'$, which has its $i$th element removed. So, we
define \var{rem\_PermFun} and \var{rem\_nth} respectively, hoping to
prove that
\begin{equation*}
  \var{apply}~(i::p)~l =
  l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l)
\end{equation*}
\lstinputlisting[
  widthgobble=4, firstline=229, lastline=229,
]{../theories/BWT/Sorting/PermFun.v}
\vspace{\parskip}
\lstinputlisting[
  widthgobble=2, firstline=599, lastline=607,
]{../theories/BWT/Lib/List.v}

@@

Because we represent the permutations as lists of outputs, we can
simply apply one permutation to the other permutation to compose them.
\lstinputlisting[
  widthgobble=2, firstline=409, lastline=409
]{../theories/BWT/Sorting/PermFun.v}
This definition, while simple, does not make it obvious that
\var{compose\_apply} holds.
\begin{equation}
  \label{eq:compose_apply}
  \var{apply}~(\var{apply}~p_2~p_1)~l = \var{apply}~p_2~(\var{apply}~p_1~l)
\end{equation}

>>>>

\acmConference
    [CPP'20]
    {ACM SIGPLAN International Conference on Certified Programs and Proofs}
    {January 20--21, 2020}{New Orleans, LA, USA}
\acmYear{2020}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM} \acmDOI{} %
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

@@

As a part of the proof of \Cref{thm:unbwt_correct}, we proved two
other noteworthy results that have not, to our knowledge, been proved
before in Coq. We implemented a least-significant-digit (LSD)
radixsort and proved its correctness,
\begin{restatable*}[radixsort\_correct]{theorem}{radixsortcorrect}
  \label{thm:radixsort_correct}%
  For all matrices $m$ with $n$ columns,
  \begin{equation*}
    \var{Sorted}~(\var{radixsort}~m~n) \land \var{StablePerm}~(\var{radixsort}~m~n)~m
  \end{equation*}
\end{restatable*}
and we proved that any two stable sorts are interchangeable,
\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
  \label{thm:stablesort_unique}
  For all functions $f, g: \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    \var{StableSort}~f \implies \var{StableSort}~g \implies
    (\forall l,f~l = g~l)
  \end{equation*}
\end{restatable*}
This in turn required developing suitable definitions for
\var{Sorted}, \var{Permutation}, and \var{StablePerm}.

@@

We represent a total, decidable preorder on type with the
\var{Preord} typeclass in Coq:
\begin{lstlisting}
Class Preord (A : Type) :=
  { le : A -> A -> Prop;
    le_trans : forall x y z, le x y -> le y z -> le x z;
    le_total : forall x y, le x y \/ le y x;
    le_dec : forall x y, {le x y} + {~le x y};
  }.
\end{lstlisting}
We do not explicitly require reflexivity because it follows from
totality. Given a \var{Preord} instance for a type $A$, we will be
able to sort $\var{list}~A$ as shown in \cref{sec:sort}.

@@

We say that we have a total, decidable \textit{order} when we have
that for all $x$ and $y$,
\begin{equation*}
  x \equiv y \implies x = y
\end{equation*}
that is, the induced equivalence relation of the preorder ($\equiv$)
implies Leibniz equality ($=$). We use the \var{Ord} class in Coq
for this:
\begin{lstlisting}
Class Ord (A : Type) `{Preord A} :=
  { eqv_eq : forall x y, eqv x y -> x = y }.
\end{lstlisting}
For the BWT, we require an \var{Ord} instance for $A$, but
elsewhere, such as in sorting, \var{Preord} will suffice.

@@

First, we define the left and right rotations of a list. \var{lrot}
moves the first element of a list to the end.
\begin{lstlisting}
Definition lrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => tl ++ [hd]
  end.
\end{lstlisting}
\var{rrot} moves the last element of a list to the front, and
relies on \var{init}, which drops the last element of a list.
\begin{lstlisting}
Fixpoint init l : list A :=
  match l with
  | [] | [_] => []
  | hd :: tl => hd :: init tl
  end.
\end{lstlisting}
\begin{lstlisting}
Definition rrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => last l hd :: init l
  end.
\end{lstlisting}
Note that in order to ensure that $\var{last}$ is total, it takes
an extra argument to return in the case of an empty list.

Now we produce the rotation matrix by repeatedly applying \var{lrot}.
We will use \var{iter}, which takes a function $f : A \to A$, a number
$n$, and an initial value $z$, and returns the list of intermediate
values obtained by applying $f$ to $z$ $n$ times. Or,
\begin{equation}
  \var{iter}~f~n~z = [z,~f~z,~f~(f~z),~\ldots,~f^{n-1} z]
  \label{eq:iter}
\end{equation}
where $f^n$ denotes repeated function application.
\begin{lstlisting}
Fixpoint iter (f : A -> A) (n : nat) (z : A)
  : list A :=
  match n with
  | O   => []
  | S m => z :: iter f m (f z)
  end.
\end{lstlisting}

We can state \Cref{eq:iter} as a theorem by defining the
function \var{rep} for repeated function application.
\begin{lstlisting}
Fixpoint rep (f : A -> A) (n : nat) (z : A) : A :=
  match n with
  | O => z
  | S m => f (rep f m z)
  end.
\end{lstlisting}
In proofs, we will still use the notation $f^i z$ for
$\var{rep}~f~i~z$. We will also use that notation $l_i$ for the $i$th
element of a list. In Coq, that is given by $\var{nth}~i~l~d$, where
$d:A$ is a default value in case $i \ge \var{length}~l$.

@@

Now we can define \var{rots} to produce the rotation matrix.
\begin{lstlisting}
Definition rots (l : list A) : list (list A) :=
  iter lrot (length l) l.
\end{lstlisting}

\var{bwl} produces the transformed string by taking the last column of
the sorted rotation matrix. \var{lexsort} sorts the rotation matrix
using insertion sort (\cref{appendix:insertion_sort}) and the induced
lexicographic ordering(\cref{sec:lex_ord}). By
\Cref{thm:stablesort_ord} though, \var{lexsort} can be implemented
using any sorting algorithm.
\begin{lstlisting}
Definition bwl (l : list A) : list A :=
  match l with
  | [] => []
  | hd :: _ =>
    List.map (fun x => last x hd) (lexsort (rots l))
  end.
\end{lstlisting}

\var{bwn} returns the index in the sorted rotation matrix where the
original string appears. It relies on \var{findIndex}, which is
defined with respect to some decidable equivalence relation.
\begin{lstlisting}
Fixpoint findIndex (x : A) (ls : list A) : nat :=
  match ls with
  | [] => 0
  | hd :: tl =>
    if x == hd then 0
    else S (findIndex x tl)
  end.
\end{lstlisting}

When we use \var{findIndex} in \var{bwn}, we are using the
induced equivalence relation from the \var{Ord} instance, so it
finds the row that is Leibniz equal to $l$.
\begin{lstlisting}
Definition bwn (l : list A) : nat :=
  findIndex l (lexsort (rots l)).
\end{lstlisting}

@@

We implement \var{recreate} as follows, introducing a parameter $j$ so
that we can recursively recreate the first $j$ columns of the matrix,
one column at a time. Here $l$ is meant to be the output of the BWT,
the last column of the sorted rotation matrix.
\begin{lstlisting}
Fixpoint recreate (j : nat) (l : list A)
  : list (list A) :=
  match j with
  | O    => map (const []) l
  | S j' => hdsort (prepend_col l (recreate j' l))
  end.
\end{lstlisting}
This definition relies on \var{prepend\_col} and \var{zipWith}, defined below.
\begin{lstlisting}
Definition prepend_col
  : list A -> list (list A) -> list (list A)
  := zipWith cons.

Fixpoint zipWith (a : list A) (b : list B)
  : list C :=
  match (a, b) with
  | (ahd :: atl, bhd :: btl) =>
    f ahd bhd :: zipWith atl btl
  | _ => []
  end.
\end{lstlisting}

Now we can redefine \var{unbwt} as
\begin{lstlisting}
Definition unbwt (i : nat) (l : list A) : list A :=
  nth i (recreate (length l) l) l.
\end{lstlisting}

Since we recreate the matrix column by column, we can define an
invariant using $\var{cols}~j$, which takes the first $j$ columns of a matrix.
\begin{lstlisting}
Definition cols (j : nat)
  : list (list A) -> list (list A) := map (firstn j).
\end{lstlisting}

@@

This leads to the following definition of the property
$\var{PermFun}~n~p$, which holds when $p : \var{list}~\mathbb{N}$
represents a permutation function on lists of length $n$:
\begin{lstlisting}
Definition PermFun (n : nat) (p : list nat)
  : Prop := NoDup p /\ forall i, In i p <-> i < n.
\end{lstlisting}
Then, we can define \var{apply}, which permutes a list according to a
given permutation:
\begin{lstlisting}
Definition apply p l : list A :=
  match l with
  | [] => []
  | d :: _ => map (fun i => nth i l d) p
  end.
\end{lstlisting}
And finally, we can define \var{PermutationEx}, which holds between
two lists when they are permutations of each other.
\begin{lstlisting}
Definition PermutationEx l l' : Prop :=
  exists p, PermFun (length l) p /\ apply p l = l'.
\end{lstlisting}
We prove \var{PermutationEx} equivalent to \var{Permutation} in
\Cref{thm:permutationex_iff}.

For completeness, we will briefly mention a third definition of a
permutation relation that says that two lists are permutations of one
another when any element occurs an equal number of times in each:
\begin{lstlisting}
Definition PermutationCount l l' : Prop := forall x,
  count_occ equiv_dec l x = count_occ equiv_dec l' x.
\end{lstlisting}
\var{count\_occ} counts the number of occurrences of a given variable
using a decision procedure for Leibniz equality. Here that decision
procedure is \texttt{equiv\_dec} from an \var{EqDec} instance where we
require the equivalence relation itself by Leibniz equality.

@@

Based on this, we define a stronger version of the \var{PermFun}
property, \var{StablePermFun}. However, while $\var{PermFun}~n~p$
means that $p$ represents a permutation function on lists of length
$n$, we need to specify the list with respect to which $p$ is stable,
so we say $\var{StablePermFun}~l~p$ for some list $l$.
\begin{lstlisting}
Definition StablePermFun (l : list A) (p : list nat)
  := PermFun (length l) p /\ forall i j d,
    nth i (apply p l) d === nth j (apply p l) d ->
    i < j < length l -> image p i < image p j.
\end{lstlisting}
As above, we can also define this
property using the inverse permutation:
\begin{lstlisting}
Definition StablePermFun_preimage
  (l : list A) (p : list nat) :=
  PermFun (length l) p /\ forall i j d,
    nth i l d === nth j l d ->
    i < j < length l -> preimage p i < preimage p j.
\end{lstlisting}

As usual, we will prove these
definitions equivalent. First, though, we need to define \var{image}
and \var{preimage}.
\begin{lstlisting}
Definition image (p : list nat) i := nth i p 0.
Definition preimage (p : list nat) i := findIndex i p.
\end{lstlisting}
We will not prove it here, but $\var{preimage}~p$ and $\var{image}~p$
are both of bijections of type $\mathbb{N}_{<n} \to \mathbb{N}_{<n}$ and
are inverses of each other. We also have the following two theorems:
\begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
  For all lists $l$, permutations $p$, and indices $i$,
  \begin{align*}
    (\var{apply}~p~l)_{p^{-1}(i)} = l_i && l_{p(i)} = (\var{apply}~p~l)_i
  \end{align*}
\end{theorem}
\begin{proof}
  The first equation (\var{nth\_preimage\_apply}) holds because:
  \begin{align*}
       (\var{apply}~p~l)_{p^{-1}(i)}
    &= (\var{map}~(\lambda j.~l_j)~p)_{\var{findIndex}~i~p} \\
    &= l_{p_{(\var{findIndex}~i~p)}} \\
    &= l_i
  \end{align*}
  and the second follows from this:
  \begin{equation*}
      (\var{apply}~p~l)_i
    = (\var{apply}~p~l)_{p^{-1}(p(i))}
    = l_{p(i)}
    \qedhere
  \end{equation*}
\end{proof}

@@

Now, we can define \var{StablePermEx} to describe when two lists are
stable permutations of each other:
\begin{lstlisting}
Definition StablePermEx l l' :=
  exists p, StablePermFun l p /\ apply p l = l'.
\end{lstlisting}

While \var{StablePermEx} corresponds to \var{PermutationEx}, how can
we modify \var{Permutation} to create an inductive relation for stable
lists? Intuitively, the problem is \var{Permutation\_swap}: we can
only swap two elements and preserve stability if they are not
equivalent. With this fix, we have \var{StablePermInd}:
\begin{gather*}
  \infer[\var{StablePermInd\_nil}]{\var{StablePermInd}~[]~[]}{}
  \\ \\
  \infer[\var{StablePermInd\_skip}]{\var{StablePermInd}~(x :: l)~(x :: l')}{%
    \var{StablePermInd}~l~l'
  }
  \\ \\
  \infer[\var{StablePermInd\_swap}]
        {\var{StablePermInd}~(x :: y :: l)~(y :: x :: l)}
        {x \not\equiv y}
  \\ \\
  \infer[\var{StablePermInd\_trans}]{\var{StablePermInd}~l~l''}{
    \var{StablePermInd}~l~l' & \var{StablePermInd}~l'~l''
  }
\end{gather*}

Unfortunately, neither of these two definitions will be particularly
useful for proving the theorems we would like to prove with them. The
best definition for stability does not come by analogy with
\var{Permutation}. The intuition is that for any two lists that are
stable permutations of each other, if we filter each list by
equivalence with any element, the resulting lists should be identical.
This is because they must have the same elements, since the original
lists are permutations of one another, and those elements must be in
the same order, since the original lists are specifically stable
permutations of one another \cite{leroy_2018}.
\begin{lstlisting}
Definition StablePerm (l l': list A) : Prop := forall x,
  filter (equiv_decb x) l = filter (equiv_decb x) l'.
\end{lstlisting}
We prove equivalence between these three definitions in
\cref{appendix:stableperm_def_eq}.

@@

If we define \var{Sort} and \var{StableSort} as we did
\cref{sec:sorted_perm_stable}, we can restate the theorem slightly
more elegantly.
\begin{lstlisting}
Definition Sort f :=
  forall l, Sorted (f l) /\ Permutation (f l) l.
\end{lstlisting}

@@

Taken together, we can define prefix lexicographic orderings on lists,
where the lexicographic ordering is only allowed to compare up to the
first $n$ elements of two lists. We define the relation $\le_n$ on
$\var{list}~A$ such that for all $xs, ys : \var{list}~A$,
\begin{equation*}
  xs \le_n ys \coloneqq (\var{firstn}~n~xs) \lexle (\var{firstn}~n~ys)
\end{equation*}
For convenience, we define $\var{PrefixSorted}~n$ to hold when a list
of lists is sorted if you only look at the first $n$ elements of each:
\begin{lstlisting}
Definition PrefixSorted (n : nat) : list (list A) -> Prop
  := Sorted (keyOrd (firstn n)).
\end{lstlisting}

@@

In a functional setting, we can define an LSD radixsort elegantly by
repeatedly moving the last column to the front of the matrix and
sorting on the first column. For a matrix with $n$ columns,
\begin{equation*}
  \var{radixsort} = (\var{hdsort} \comp \var{map}~\var{rrot})^n
\end{equation*}
Or, in Coq:
\begin{lstlisting}
Definition radixsort m (n : nat) : list (list A)
  := rep (hdsort ∘ map rrot) n m.
\end{lstlisting}
Here \var{hdsort} is just an insertion sort using $\le_1$ ordering.

@@

\begin{lstlisting}
  Fixpoint insert (x : A) (l : list A) :=
    match l with
    | [] => [x]
    | h :: t =>
      if le_dec x h then x :: h :: t else h :: insert x t
    end.

  Fixpoint sort (l : list A) : list A :=
    match l with
    | [] => []
    | h :: t => insert h (sort t)
    end.
\end{lstlisting}

@@

We know we will eventually need to reason inductively about
permutation functions, so we would like to define the relationship
between applying a permutation function $i :: p$ and applying $p$.
Conceptually, the first element of $\var{apply}~(i::p)~l$ will be
$l_i$. The rest will be the result of applying some new permutation
$p'$, where everything greater than $i$ has been shifted down by one,
to some new list $l'$, which has its $i$th element removed. So, we
define \var{rem\_PermFun} and \var{rem\_nth} respectively, hoping to
prove that
\begin{equation*}
  \var{apply}~(i::p)~l =
  l_i :: \var{apply}~(\var{rem\_PermFun}~i~p)~(\var{rem\_nth}~i~l)
\end{equation*}
\begin{lstlisting}
Definition rem_PermFun i :=
  map (fun j => if lt_dec i j then pred j else j).

Fixpoint rem_nth i l :=
  match l with
  | [] => []
  | h :: t =>
    match i with
    | 0 => t
    | S i' => h :: rem_nth i' t
    end
  end.
\end{lstlisting}

@@

Because we represent the permutations as lists of outputs, we can
simply apply one permutation to the other permutation to compose them.
\begin{lstlisting}
Definition compose p2 p1 := apply p2 p1.
\end{lstlisting}
This definition, while simple, does not make it obvious that
\var{compose\_apply} holds.
\begin{equation}
  \label{eq:compose_apply}
  \var{apply}~(\var{apply}~p_2~p_1)~l = \var{apply}~p_2~(\var{apply}~p_1~l)
\end{equation}
